---
title: "Homework 2"
author: "Jiarui Fu"
date: "10/4/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
```

## Problem 1
This problem uses the Mr. Trash Wheel dataset.  
  Read and clean the Mr. Trash Wheel sheet:
```{r Mr.Trash Wheel, message = FALSE}
# read and clean the Mr. Trash Wheel sheet
trashwheel_data = 
  # specify the sheet
  # omit non-data entries (figures & notes)
  read_excel("data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = 1, skip = 1) %>%
  janitor::clean_names() %>%
  select(dumpster:homes_powered) %>%
  # omit rows that do not include dumpster-specific data
  drop_na(dumpster) %>%
  # round the number of sports balls to the nearest integer
  # convert the result to an integer variable
  mutate(sports_balls = as.integer(round(sports_balls, digits = 0)))

trashwheel_data
```

For the Mr. Trash Wheel sheet (clean), the size of the dataset is `r nrow(trashwheel_data)` rows and `r ncol(trashwheel_data)` columns.  
  The key variables are: dumpster<dbl>, month<chr>, year<dbl>, date, weight_tons<dbl>, volume_cubic_yards<dbl>, plastic_bottles<dbl>, polystryene<dbl>, cigarette_butts<dbl>, glass_bottles<dbl>, grocery_bags<dbl>, chip_bags<dbl>, spots_balls<int> and homes_powered<dbl>.

Read and clean precipitation data for 2017 and 2018:
```{r Precipitation for 2017 and 2018, message = FALSE}
# read and clean precipitation data for 2017
precipitation_2017 = 
  # omit rows without precipitation data
  read_excel("data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = 6,  skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(month) %>% 
  # add a variable year
  mutate(year = 2017) %>% 
  # convert month to a character variable
  mutate(month = month.name)

precipitation_2017

# read and clean precipitation data for 2018
precipitation_2018 = 
  # omit rows without precipitation data
  read_excel("data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = 5,  skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(month) %>% 
  # add a variable year
  mutate(year = 2018) %>% 
  # convert month to a character variable
  mutate(month = month.name)

precipitation_2018

# combine precipitation datasets
combined_precipitation = 
  rbind(precipitation_2017, precipitation_2018) 

combined_precipitation

# create a new dataset that only includes the information of year 2017 in the trashwheel sheet
trashwheel_2017 = filter(trashwheel_data, year == 2017)
```

For the 2017 precipitation dataset (clean), the size is `r nrow(precipitation_2017)` rows and `r ncol(precipitation_2017)` columns.  
  The key variables are: month, total and year.

For the 2018 precipitation dataset (clean), the size is `r nrow(precipitation_2018)` rows and `r ncol(precipitation_2018)` columns.  
  The key variables are: month, total and year.

For the combined precipitation dataset, the size is `r nrow(combined_precipitation)` rows and `r ncol(combined_precipitation)` columns.  
  The key variables are: month, total and year.
  
The total precipitation in 2018 was `r sum(pull(precipitation_2018, total))`.  
  The median number of sports balls in a dumpster in 2017 was `r median(pull(trashwheel_2017, sports_balls))`.

## Problem 2
This problem uses the FiveThirtyEight data.  
  Clean the data in pols-month.csv, snp.csv, and tidy the unemployment data so that it can be merged with the previous datasets: 
```{r FiveThirtyEight, message = FALSE}
# clean the data in pols-month.csv
pols_month = 
  read_csv("data/pols-month.csv") %>%
  janitor::clean_names() %>% 
  # modify the mon info so that the year format is 4-digit
  mutate(mon = lubridate::ymd(mon)) %>% 
  # break up the variable mon into variables year, month and day
  separate("mon", sep = "-", into = c("year", "month", "day")) %>% 
  # convert variables year, month and day into integers
  mutate(year = as.integer(year),
         month = as.integer(month),
         day = as.integer(day)) %>% 
  # replace month number with month name
  mutate(month = month.abb[month]) %>%
  # create a president variable taking values of gop and dem
  # recode the prez_gop variable
  mutate(president = recode(prez_gop, "0" = "Democrat", "1" = "Republican", "2" = "Republican")) %>% 
  # remove prez_gop, prez_dem and the day variable
  select(-prez_gop, -prez_dem, -day) %>% 
  # organize so that year and month are the leading columns
  select(year, month, everything()) 
pols_month

# clean the data in snp.csv
snp = 
  read_csv("data/snp.csv") %>% 
  janitor::clean_names() %>% 
  # break up the variable mon into variables year, month and day
  separate("date", sep = "/", into = c("month", "day", "year")) %>% 
  # convert variables year, month and day into integers
  mutate(month = as.integer(month),
         day = as.integer(day),
         year = as.integer(year)) %>% 
   # arrange according to year and month
  arrange(year, month) %>% 
  # replace month number with month name
  mutate(month = month.abb[month]) %>%
  # remove the day variable
  select(-day) %>% 
  # organize so that year and month are the leading columns
  select(year, month, everything())
snp

# tidy the unemployment data
unemployment = 
  read_csv("data/unemployment.csv") %>%
  # switch from wide to long format
  pivot_longer(Jan:Dec,
               names_to = "month",
               values_to = "unemployment") %>% 
  # ensure that key variables have the same name and take the same values
  rename(year = Year) %>% 
  mutate(year = as.integer(year))
unemployment

# join the datasets
# merge snp into pols
pols_snp = full_join(pols_month, snp, by = c("month", "year")) 
pols_snp

# merge unemployment into pols_snp
final_dataset = full_join(pols_snp, unemployment, by = c("month", "year"))
final_dataset
```

The pols-mon.csv has 822 rows and 9 columns; the range of years is from 1947 to 2015; names of the key variables are year, month, gov_gop, sen_gop, rep_gop, gov_dem, sen_dem, rep_dem and president.

The snp.csv has 787 rows and 3 columns; the range of years is from 1950 to 2015; names of the key variables are year, month and close.

The unemployment.csv has 816 rows and 3 columns; the range of years is from 1948 to 2015; names of the key variables are year, month and unemployment.

The (final) merged dataset of all three datasets mentioned above has 828 rows and 11 columns; the range of years is from 1947 to 2015; names of the key variables are year, month, gov_gop, sen_gop, rep_gop, gov_dem, sen_dem, rep_dem, president, close and unemployment.

## Problem 3
This problem uses data from NYC Open data on the popularity of baby names.
``` {r popular baby names, message = FALSE}
# load and tidy the data
popular_baby_names = 
  read_csv("data/Popular_Baby_Names.csv") %>%
  janitor::clean_names() %>% 
  # re-format the case structure of string variables - capitalized 'Name'
  mutate(childs_first_name = str_to_title(childs_first_name)) %>% 
  # re-name the name of categorical predictor - abbrevation to full spelling
  mutate(ethnicity = recode(ethnicity, "ASIAN AND PACI" = "ASIAN AND PACIFIC ISLANDER", 
                                       "BLACK NON HISP" = "BLACK NON HISPANIC",
                                       "WHITE NON HISP" = "WHITE NON HISPANIC")) %>% 
  # remove duplicated rows
  distinct()

popular_baby_names

# create tables
# show the rank in popularity of the name "Olivia" as a female baby name over time
# create a new data frame
data_Olivia = filter(popular_baby_names, childs_first_name == "Olivia") 

# rows for ethnicities and columns for years
table_Olivia = pivot_wider(data_Olivia,
                           id_cols = c(year_of_birth, ethnicity, rank),
                           names_from = year_of_birth,
                           values_from = rank)

# generate table
knitr::kable(table_Olivia, 
             caption = "Table: The rank in popularity of the name Olivia among female children over time")

# show the most popular name among male children over time
# create a new data frame
data_male_names = filter(popular_baby_names, gender == "MALE", rank == 1)

# rows for ethinicities and columns for years
table_male_names = pivot_wider(data_male_names,
                               id_cols = c(year_of_birth, ethnicity, childs_first_name),
                               names_from = year_of_birth,
                               values_from = childs_first_name)

# generate table
knitr::kable(table_male_names, caption = "Table: The most popular name among male children over time")

# produce a scatter plot showing the number of children with a name (y) against the rank in popularity of that name (x)
# create a new data frame
data_white_male = filter(popular_baby_names, 
                         gender == "MALE", 
                         ethnicity == "WHITE NON HISPANIC", 
                         year_of_birth == 2016)

# produce scatter plot
# legend is not shown due to the sheer size of the names
ggplot(data_white_male, aes(x = rank, y = count)) + geom_point() + 
       ggtitle("Number of Children with a Name against Rank in Popularity of the Name") + 
       xlab("The rank in popularity of a name") + ylab("The number of children with a name") 
```















